---
title: "HW 5"
author: "Joe Brew"
date: "10/05/2014"
output: html_document
---

The code for the production of this document is available at <https://github.com/joebrew/uf>.

# Task 1
## For each family address, please find the nearest EMS and calculate the straight-line distance.  List the family address ID and EMS ID and the nearest distance below.  
(If you use the "Near" tool, please listthe family address ID "OBJECTID," the EMS ID "NEAR_FID," and the nearest distance "NEAR_DIST."" If you use "Generate Near Table," Please list "IN_FID," "NEAR_FID" and "NEAR_DIST").

Set working directory:
```{r}
if ( Sys.info()["sysname"] == "Linux" ){
  setwd("/home/joebrew/")
} else {
  setwd("C:/Users/BrewJR/")
}

mywd <- paste0(getwd(), "/Documents/uf/phc6194/hw5")
setwd(mywd)
```

Once could read in the mdb file using the Hmisc library and mdb-tools.
```{r, echo = TRUE, eval = FALSE, message = FALSE}
# Attach the necessary package
library(Hmisc)

# View which tables are available
mdb.get("Homwork5.mdb", tables = TRUE)

# Read in the Family_address table
fam <- mdb.get("Homwork5.mdb", tables = "Family_address")
gdb <- mdb.get("Homwork5.mdb", tables = "GDB_Items")
road <- mdb.get("Homwork5.mdb", tables = "road")
```

However, the above method doesn't (easily) keep the spatial aspects associated with each table.  Insead, I used ArcGIS to read in the .mdb database, and then exported the road, EMS and addresses shapefiles seperately.  

```{r, message = FALSE, eval = TRUE, echo = FALSE}
# library(rgdal)
# road <- readOGR(".", "road")
# ems <- readOGR(".", "EMS_LOCATION")
# fam <- readOGR(".", "Family_address")
# save.image("hw5temp.RData")
library(rgdal)
load("hw5temp.RData")
```


```{r, message = FALSE, eval = FALSE, echo = TRUE}
library(rgdal)
road <- readOGR(".", "road")
ems <- readOGR(".", "EMS_LOCATION")
fam <- readOGR(".", "Family_address")
```

Check that our projection and coordinate systems are identical:
```{r}
proj4string(ems) == proj4string(fam)
```

Confirm that we read everything in okay by plotting our map and points
```{r}
plot(road, col = adjustcolor("black", alpha.f = 0.6))
points(ems, col = adjustcolor("red", alpha.f = 0.6), pch = 16)
points(fam, col = adjustcolor("blue", alpha.f = 0.6), pch = 16)
legend("topright", pch = 16, col = adjustcolor(c("red", "blue"), alpha.f = 0.6),
       legend = c("EMS locations", "Homes"))

```

Check what kind of unit of measurement we're working with
```{r}
proj4string(ems) # US-FT
```


Merge our EMS and family points into one dataframe.  
```{r}
fam_coords <- data.frame(cbind(coordinates(fam), group = "fam"), stringsAsFactors = FALSE)
ems_coords <- data.frame(cbind(coordinates(ems), group = "ems"), stringsAsFactors = FALSE)

# Create a list of the two collections of coordinates
x <- list(ems_coords, fam_coords)
# Bind all elementsin the list together
x <- do.call("rbind", x)
# Now we have x, a dataframe of the coordinates

# Make it spatial
x$coords.x1 <- as.numeric(x$coords.x1)
x$coords.x2 <- as.numeric(x$coords.x2)
coordinates(x) <- ~ coords.x1 + coords.x2
```

Replace ems/fam_coords with their spatial equivalents
```{r}
ems_coords <- x[which(x$group == "ems"),]
fam_coords <- x[which(x$group == "fam"),]
```

Write a function for calculating distance (in feet) using pythagorean theorem
```{r}
DistFun <- function(x, y, x2, y2){
  xdist <- sqrt((x - x2)^2)
  ydist <- sqrt((y - y2)^2)
  
  linedist <- sqrt((xdist^2) + (ydist^2))
  return(linedist)
}
```

Now loop that function over each address fam_coords to extract the index of the closest ems
```{r}
fam_coords$closest_ems_name <- vector(mode = "numeric", length = nrow(fam_coords))
fam_coords$closest_ems_distance <- vector(mode = "numeric", length = nrow(fam_coords))
fam_coords$closest_ems_index <- vector(mode = "numeric", length = nrow(fam_coords))

for (i in 1:nrow(fam_coords)){
  # Get the distance of every ems station from every address
  temp <- DistFun(x = coordinates(fam_coords)[i,1],
        y = coordinates(fam_coords)[i,2],
        x2 = coordinates(ems)[,1], 
        y2 = coordinates(ems)[,2])
  # Extract the index number of the closest ems station
  best.ind <- which.min(temp)  
  
  #Assign that index and distance to fam_coords
  fam_coords$closest_ems_index[i] <- best.ind
  fam_coords$closest_ems_name[i] <- as.character(ems$DESCRIPT[best.ind])
  fam_coords$closest_ems_distance[i] <- temp[which.min(temp)]

}

# Just for fun, let's make a miles column as well
fam_coords$closest_ems_miles <- fam_coords$closest_ems_distance / 5280

# Throw back in the object id
fam_coords$OBJECTID <- 1:nrow(fam_coords)
fam_coords$address <- fam$FULLADDR
```

Show the table with the closest EMS stations for each house.
```{r, results = "asis"}
library(knitr)
z <- data.frame(fam_coords[,c("OBJECTID", 
                         "address", "closest_ems_name", 
                         "closest_ems_miles", 
                         "closest_ems_index")])
# print(xtable(z), type = "html")
kable(z, format = "markdown", row.names = NA)
```


# Task 2
## Use network analysis to estimate the shortest response time of an EMS (emergence medical service) to the family addresses following the in- class tutorials.

Steps:
1. Data Preparation: Assign travel time to road segment following the in- class tutorials;
2. Build Network Dataset;
3. Estimate shortest travel time
4. Note: Select "new OD Cost Matrix" from Network Analyst and Select "Family_house" as Origins and "EMS_Location" as Destinations.  Please give a good screen shot to show your result.

I took an alternative approach to this problem.  Using the ggmap package, I calculated the *true* travel time from every address to every EMS location.  The advantage of this approach is that it takes into account travel times from Google's API, rather than simply calculating speed limit (which likely underestimates true travel time).

First, I attach the ggmap library.
```{r}
library(ggmap)
```

Now, let's get latitude and longitude for each point (so as to more smoothly interact with google's API):
```{r}
#Get a projection string for both
proj4string(ems_coords) <- proj4string(ems)
proj4string(fam_coords) <- proj4string(ems)

#Convert to latitude / longitude
ems_latlon <- spTransform(ems_coords, CRS("+init=epsg:4326"))
fam_latlon <- spTransform(fam_coords, CRS("+init=epsg:4326"))

```

Then I write a function to wrap around ggmap's mapdist() function
```{r}
TravelTime <- function(from, to){
  mapdist(from, to, mode = "driving")
}
```

Finally, I loop my TravelTime function over each row of the address points.
```{r, eval = FALSE, echo = TRUE}
# Create empty columns for our new variables
fam_coords$closest_ems_travel_time_minutes <- vector(mode = "numeric", length = nrow(fam_coords))
fam_coords$closest_ems_travel_distance_km <- vector(mode = "numeric", length = nrow(fam_coords))
fam_coords$closest_ems_travel_index <- vector(mode = "numeric", length = nrow(fam_coords))
fam_coords$closest_ems_travel_name <- vector(mode = "numeric", length = nrow(fam_coords))

#Loop through each row of family addresses, calculating distance/time from each ems
for (i in 1:nrow(fam_coords)){
  tofam <- paste0(coordinates(fam_latlon)[,2], ", ",coordinates(fam_latlon)[,1])[i]
  fromems <- paste0(coordinates(ems_latlon)[,2], ", ",coordinates(ems_latlon)[,1])
  
  x <- TravelTime(from = fromems,
             to = tofam)
  
 # Assing the index of the best (least time) ems station 
  best.ind <- which.min(x$minutes)[1]
 
 # Put info back into fam_coords
fam_coords$closest_ems_travel_time_minutes[i] <- x$minutes[best.ind]
fam_coords$closest_ems_travel_distance_km[i] <- x$km[best.ind]
fam_coords$closest_ems_travel_index[i] <- best.ind
fam_coords$closest_ems_travel_name[i] <- as.character(ems$DESCRIPT[best.ind])

}
```

```{r, eval = TRUE, echo = FALSE}
#save.image("post_google.RData")
load("post_google.RData")
fam_coords$closest_ems_travel_time_minutes <- fam_coords$closest_ems_travel_time
fam_coords$closest_ems_travel_time <- NULL
```
  
Show the table with the closest EMS stations for each house.
```{r, results = "asis"}
library(knitr)
z <- data.frame(fam_coords[,c("OBJECTID", 
                         "address", "closest_ems_travel_name", 
                         "closest_ems_travel_distance_km", 
                         "closest_ems_travel_time_minutes",
                         "closest_ems_index")])
# print(xtable(z), type = "html")
kable(z, format = "markdown", row.names = NA)
```

Note that there is a correlation between distance and time, but it's not perfectly linear.
```{r}
plot(fam_coords$closest_ems_travel_distance_km,
     fam_coords$closest_ems_travel_time_minutes,
     xlab = "Distance (in kilometers)",
     ylab = "Minutes to arrive",
     pch = 16,
     col = adjustcolor("darkgreen", alpha.f = 0.6),
     cex = 2)

  #ADD LOESS LINE
  lox <- fam_coords$closest_ems_travel_distance_km
  loy <- fam_coords$closest_ems_travel_time_minutes
  lw1 <- loess(loy ~ lox, span=1)
  j <- order(lox)
  lines(lox[j],lw1$fitted[j],col=adjustcolor("red", alpha.f = 0.7),
        lty=1,
        lwd = 2)

# ADD Regression line
fit <- lm(loy ~ lox)
abline(fit, col = adjustcolor("blue", alpha.f = 0.7),
       lwd = 2)
legend("bottomright",
       col = adjustcolor(c("red", "blue"), alpha.f = 0.7),
       lty = 1,
       lwd = 2,
       legend = c("Loess", "Least squares"),
       title = "Lines of best fit")

```

Just for fun, let's plot the family addresses with their color being a function of the time they are from an EMS response (red being slow, blue being fast).

```{r}
library(RColorBrewer)
my_colors <- colorRampPalette(c("darkblue", "darkred"))(ceiling(max(fam_coords$closest_ems_travel_time_minutes)))
fam_coords$color <- my_colors[ceiling(fam_coords$closest_ems_travel_time_minutes)]

plot(road, col = adjustcolor("black", alpha.f = 0.6))
points(ems, col = adjustcolor("darkgreen", alpha.f = 0.5), pch = 17)
points(fam, col = adjustcolor(fam_coords$color, alpha.f = 0.9), pch = 16, cex = 2)
legend("topright", pch = c(17,16), col = adjustcolor(c("darkgreen", "purple"), alpha.f = 0.6),
       legend = c("EMS locations", "Homes"))
legend("left", fill = my_colors, bty = "n",
       y.intersp = 0.5, ncol=1, legend=c(1, rep(NA, 4), 6, rep(NA, 4), 11),
       border = NA, title = "Response time (minutes)")

```

And finally, an interactive map!
```{r, results = "asis"}

library(rCharts)

# read in geojson version of zip
#zip_geoj <- readOGR("zips_alachua", "ACDPS_zipcode")
mymap <- Leaflet$new()
mymap$tileLayer(provider = "Stamen.TonerLite")
mymap$setView(c(29.65, -82.3), zoom = 10)
mymap$enablePopover(TRUE)


library(rCharts)
library(rMaps)
mywaypoints = list(c(40.74119, -73.9925), c(40.73573, -73.99302))


mywaypoints <- list()
for (i in 1:nrow(fam_coords)){
  mywaypoints[[i]] <- (c(as.numeric(coordinates(fam_latlon)[i,2]),
                 as.numeric(coordinates(fam_latlon)[i,1])))
  
}


mymap$addAssets(
  css = "http://www.liedman.net/leaflet-routing-machine/dist/leaflet-routing-machine.css",
  jshead = "http://www.liedman.net/leaflet-routing-machine/dist/leaflet-routing-machine.min.js"
)

routingTemplate = "
 <script>
 var mywaypoints = %s
 L.Routing.control({
  waypoints: [
    L.latLng.apply(null, mywaypoints[0]),
    L.latLng.apply(null, mywaypoints[1])
  ]
 }).addTo(mymap);
 </script>"

mymap$setTemplate(
  afterScript = sprintf(routingTemplate, RJSONIO::toJSON(mywaypoints))
)
mymap$set(width = 1450, height = 800)

mymap

#mymap$fullScreen(TRUE)

# for (i in 1:nrow(fam_coords)){
#   mymap$marker(c(as.numeric(coordinates(fam_latlon)[i,2]),
#                  as.numeric(coordinates(fam_latlon)[i,1])),
#                bindPopup = paste(round(fam_coords$closest_ems_travel_time_minutes[i], digits=1), "minutes for", 
#                                  fam_coords$closest_ems_travel_name[i], "to respond to",
#                                  fam_coords$address[i]
#                                  ))
#   
# }

#library(shiny)
# renderMap(mymap)



```
